Parallel Adaptive Importance Sampling

### 2015-08-25 PAIS.tex
some final edits in the conclusions mostly.

\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
to use MCMC methods for many more. In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section
\ref{sec:filters}, and the pCNL proposal distribution described in
Section \ref{Sec:pCNL}

@@

Important samplers can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen from which we
can draw samples, and each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

@@

Importantly, we have compared the efficiency of our parallel scheme
with a naive parallelisation of serial methods. Thus our increase in
efficiency is over and above an $N$-fold increase, where $N$ is the
number of cores or processors at our disposal. Our approach
demonstrates a better-than-linear speed-up with the number
processors/cores used. Thus, our approach is not only embarrassingly
parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
humiliatingly so.

@@

The PAIS approach is limited by the curse of dimensionality, since the
cost of the resampling algorithm becomes prohibitively large. The cost
of the resampler is $\mathcal{O}(n^2)$, where $n$ is the number of
particles in the ensemble. Other resamplers exist which are not as
costly, but for which there is some sacrifice in terms of optimality
of the output. For example, the resampler proposed in
\cite{schefzik2013uncertainty}  has computational complexity
$\mathcal{O}(n)$. This might allow the PAIS to be used in higher
dimensional problems, and is an avenue for future investigation.

@@

>>>>

\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and
with the development of and with the development of faster more
powerful computers, have become ever more important in a whole range
of fields in statistics, science and engineering.  In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section
\ref{sec:filters}, and the pCNL proposal distribution described in
Section \ref{Sec:pCNL}.

@@

Important sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use the states of a
set of parallel MCMC chains to construct a proposal distribution which
will be as close as possible to the target density. Given enough
processors, the states of all of these chains at one point
in time may be reasonably representative sample of the target density.

@@

Importantly, we have compared the efficiency of our parallel scheme
with a naive parallelisation of serial methods. Thus our increase in
efficiency is over and above an $N$-fold increase, where $N$ is the
number of cores or processors at our disposal. Our approach
demonstrates a better-than-linear speed-up with the number
processors/cores used. Thus, our approach is not only embarrassingly
parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
humiliatingly so, provided that an optimal transport algorithm can be
efficiently implemented in parallel and so does not dominate the
communication costs.

@@

@@

One limitation of the PAIS approach as described above is that a
direct solver (such as FastEMD \cite{FastEMD}) has computational cost
$\mathcal{O}(n^3\log n)$, where $n$ is the number of particles in the
ensemble.  The use of approximate solutions from iterative solvers
might allow the PAIS to be used in higher dimensional problems, and is
an avenue for future investigation. A second limitation is associated
with the problem of accurately approximating measures with empirical
measures in high dimensional phase space. This is a well-known issue
in filtering, where one possible solution is to ``localise'' the
impact of observations, as rigorously analysed in
\cite{rebeschini2013can}. An established technique in Ensemble Kalman
filters, localisation has also been incorporated into the Ensemble
Transform Particle Filter \cite{ChRe2015}. This issue could be
addressed in the PAIS by using localisation when computing the
transform, but keeping the unlocalised weights in the importance
sampling for consistency.
